Early Warning Signals and the Prosecutor's Fallacy

### 2012-06-19 fallacy.tex
finishing up corrections

\documentclass[authoryear,5p]{elsarticle}
%\documentclass[authoryear,review,12pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
%\linenumbers
\usepackage{float}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}
\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

@@

The prosecution has only observed that the probability of seeing the
evidence ($E$) they produced given a random innocent individual ($I$),
$P(E|I)$ is very small.  From this one cannot conclude that the individual is indeed guilty, that is, that the
probability the individual is innocent given the evidence $P(I|E)$
is also very small. In a city with millions of people, there might
be several individuals who match the description of the evidence.
\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$, instead,
these expressions are related by Bayes theorem,

@@

} The California
Supreme Court reversed the decision, and the case became a widely
recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
Here we explore how a similar misconception can arise from the
use of historical data to evaluate methods for detecting early warning
signals of critical transitions.

@@

Catastrophic transitions or tipping points, where a complex system
shifts suddenly from one state to another, have been implicated in
a wide array of ecological and global climate systems such as lake
ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
savannah~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and tropical
forests~\citep{Hirota2011}.  Recent research has begun to identify
statistical patterns commonly associated with these sudden catastrophic
transitions which could be used as an \emph{early warning sign} to
identify an approaching tipping point, which might provide managers time
to react to and avert an undesirable state shift~\citep{Scheffer2009, Lenton2011}.
An array of statistical patterns associated with tipping
point phenomena has been suggested for the detection of early warning
signals associated with such sudden transitions.  Two of the most commonly
used are a pattern of increasing variance~\citep{Carpenter2006} and a
pattern of increasing autocorrelation~\citep{vanNes2007}, which have
been tested in both experimental manipulation~\citep{Drake2010,
Carpenter2011, Veraart2011} and historical
observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.

@@

\begin{figure}
  \begin{center}
    \subfigure[\textbf{True criminal} A bifurcation
    drives a transition into the undesirable state]{
    \includegraphics[width=.4\textwidth]{bifurcation.pdf}
    \label{fig:bifurcation} } \subfigure[\textbf{Innocent suspect} A
    transition to the alternative stable state has occurred purely by
    chance.]{ \includegraphics[width=.4\textwidth]{fallacytransition.pdf}
    \label{fig:fallacy} }
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
  state vs potential in two cases.  Potential is the negative integral
  of the mean of dynamical equation.  The potential function gives an
  intuitive picture of the stability of a system by imagining the curve as
  a surface on which a ball is free to bounce across -- wells correspond
  to stable points and peaks to unstable points.  Early warning signals
  are aimed at detecting systems which are slowly moving towards a tipping
  point or bifurcation, illustrated in the successively darker potential
  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
  a simulation under this process shows the state of the system as the
  potential moves towards the bifurcation point. The state of the system
  in the simulation is given by the horizontal axis, while the vertical
  dimension is used to show time on the right-hand axis.  Some stable
  systems, panel~\subref{fig:fallacy}, may experience chance transitions
  to an alternate stable state without the presence of this slow change
  driving the system over the edge.  An example of the trajectory of
  such a chance transition is plotted over the potential.  While such
  events may be unlikely over short time-scales, selecting historical
  examples of systems that have experienced sudden transitions increases
  the probability of accidentally using chance events depicted in
  panel~\subref{fig:fallacy} to test methods designed to detect the kind
  of process depicted in panel~\subref{fig:bifurcation}.} \label{fig:1}
\end{figure}

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulating a system that is not driven towards a bifurcation (such as
in the right-hand panel).  This simulation approach allows us to determine whether
examining historical events is a good way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
    b_n &= \frac{e K n^2}{n^2 + h^2} \\
    d_n &= e n + a
\end{align}

a model with a linear death rate and density-dependent birth
rate that drives the Allee effect at low densities and limits
growth at high densities.  The key feature of this model is the
alternate stable states introduced by this effect; other functional
forms for Eq.~\eqref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
system can be forced through a bifurcation by increasing
the death rate, in these simulations all parameters are held constant
and no bifurcation occurs.  Consequently we do not anticipate an early
warning signal of an approaching bifurcation.  The simulation starts
from the positive equilibrium population size.  Though the chance of
a transition across the Allee threshold in any given time step is 
small, given enough time this system will eventually experience such a
rare event driving the population extinct.  We ran each replicate over
50,000 time units, sampling the system every 50 time units.  In this
time window 266 of the 1,000 replicates experience population collapse.
To keep the examples of comparable sample size, we focus on a section of
the data 500 time points prior to the system approaching the transition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=6in]{indicators.pdf}
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in gray bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem embodied is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases from some larger historical sample.  
%We note as well that the Prosecutors Fallacy 
%may arise in other contexts where we search historical data for
%examples containing some feature of interest.

>>>>

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,12pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}
\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

@@

The prosecution has only observed that the probability of seeing the
evidence ($E$) they produced given a random innocent individual ($I$),
$P(E|I)$ is very small.  From this one cannot conclude that the individual
is indeed guilty, that is, that the probability the individual is innocent
given the evidence $P(I|E)$ is also very small. In a city with millions
of people, there might be several individuals who match the description
of the evidence\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$,
instead, these expressions are related by Bayes theorem,

@@

}. The California Supreme Court reversed the decision, and
the case became a widely recognized example of the Prosecutor's
Fallacy~\citep{Thompson1987}.  Here we explore how a similar misconception
can arise from the use of historical data to evaluate methods for
detecting early warning signals of critical transitions.

@@

Catastrophic transitions or tipping points, where a complex system
shifts suddenly from one state to another, have been implicated in
a wide array of ecological and global climate systems such as lake
ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
savannah~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and tropical
forests~\citep{Hirota2011}.  Recent research has begun to identify
statistical patterns commonly associated with these sudden catastrophic
transitions which could be used as an \emph{early warning sign} to
identify an approaching tipping point, which might provide managers time
to react to and avert an undesirable state shift~\citep{Scheffer2009, Lenton2011}.
An array of statistical patterns associated with tipping
point phenomena has been suggested for the detection of early warning
signals associated with such sudden transitions.  Two of the most commonly
used are a pattern of increasing variance~\citep{Carpenter2006} and a
pattern of increasing autocorrelation~\citep{vanNes2007}, which have
been tested in both experimental manipulation~\citep{Drake2010,
Carpenter2011, Veraart2011, Dai2012} and historical
observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.

@@

\begin{figure}
  \begin{center}
    \subfigure[\textbf{True criminal} A bifurcation
    drives a transition into the undesirable state]{
    \includegraphics[width=.4\textwidth]{bifurcation.pdf}
    \label{fig:bifurcation} } \subfigure[\textbf{Innocent suspect} A
    transition to the alternative stable state has occurred purely by
    chance.]{ \includegraphics[width=.4\textwidth]{fallacytransition.pdf}
    \label{fig:fallacy} }
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
  state versus potential in two cases.  Potential is the negative integral
  of the mean of dynamical equation.  The potential function gives an
  intuitive picture of the stability of a system by imagining the curve as
  a surface on which a ball is free to bounce across -- wells correspond
  to stable points and peaks to unstable points.  Early warning signals
  are aimed at detecting systems which are slowly moving towards a tipping
  point or bifurcation, illustrated in the successively darker potential
  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
  a simulation under this process shows the state of the system as the
  potential moves towards the bifurcation point. The state of the system
  in the simulation is given by the horizontal axis, while the vertical
  dimension is used to show time on the right-hand axis.  Some stable
  systems, panel~\subref{fig:fallacy}, may experience chance transitions
  to an alternate stable state without the presence of this slow change
  in the shape of the potential well which tips the system off to the left.
  An example of the trajectory of such a chance transition is
  plotted over the potential.  While such events may be unlikely over
  short time-scales, selecting historical examples of systems that
  have experienced sudden transitions increases the probability of
  accidentally using chance events depicted in panel~\subref{fig:fallacy}
  to test methods designed to detect the kind of process depicted in
  panel~\subref{fig:bifurcation}.} \label{fig:1}
\end{figure}

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulating a system that is not driven towards a bifurcation such as
in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
examining historical events is a valid way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
    d_n &= e n + a,
\end{align}

a model with a linear death rate and density-dependent birth
rate that drives the Allee effect at low densities and limits
growth at high densities.  In this model $n$ indicates the discrete
number of individuals in the population, $K$ indicates a carrying 
capacity as set by a limiting resource, $e$ a per-capita death rate, 
$a$ an additional mortality imposed on the population such as harvest,
$h$ is a parameter controlling at what population size the addition of 
more individuals switches from conferring a positive benefit on growth 
from Allee interactions $n < h$ to a negative impact on growth due to 
increased competition, $n > h$.  The key feature of this model is the
alternate stable states introduced by this effect; other functional
forms for Eq.~\eqref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
system can be forced through a bifurcation by increasing
the death rate, in these simulations all parameters are held constant
and no bifurcation occurs.  Consequently we do not anticipate an early
warning signal of an approaching bifurcation.

The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the Allee threshold in any
given time step is small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We ran
each replicate over 50,000 time units, sampling the system every 50
time units.  In this time window 266 of the 1,000 replicates experience
population collapse.  To keep the examples of comparable sample size,
we focus on a section of the data 500 time points prior to the system
approaching the transition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=6in]{indicators.pdf}
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem embodied is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases from some larger historical sample.  
%We note as well that the Prosecutors Fallacy 
%may arise in other contexts where we search historical data for
%examples containing some feature of interest.
